### 2017-01-25 manuscript.rmd
more polishing, added some references

A rich theoretical and statistical methodology of phylogenetic comparative methods have been developed over the last three decades to address the challenges and opportunities of comparisons across species [@Felsenstein:1985ua; @Grafen:1989um; @Chamberlain:2012vx; @Garamszegi:2014ew; @Pagel:1999gc]. These challenges include non-independence of obserbations across species -- more closely related species will tend to be more similar to each other than distantly related species. Phylogenetic comparative methods have largely been applied to the evolution of morphological and ecological traits, but are just as relevant to CFG. Most CFG studies have abstained from phylogenetic approaches and instead heavily rely on multiple pairwise comparisons (Fig. 1A) between contemporary traits that do not account for the fact that evolutionary relationships explain much of the structure of variation across species. This leaves CFG studies susceptible to serious statistical errors and is a missed opportunity to ask questions that are only accessible in an explicit phylogenetic context.

@@

The KMRR study [@KryuchkovaMostacci:2016iw] sought to test the ortholog conjecture. The ortholog conjecture [@Nehrt:2011bm] is the proposition that orthologs (genes that diverged from each other due to a speciation event) have more similar attributes than do paralogs (genes that diverged from each other due to a gene duplication event). The ortholog conjecture has important biological and technical implications. It shapes our understanding of the functional diversity of gene families. It is used to relate findings from well-studied genes to related genes that have not been investigated in detail. It can be applied to any trait of genes, from gene seqeunce to biochemical properties to expression. While the ortholog conjecture describes a specific pattern of functional diversity across genes, it is also articulated as a hypothesis about the process of evolution-- that there is greater evolutionary change in gene attributes following a duplication event than a speciation event.

@@

The independent contrast across a node is the amount of change observed between the daughter notes, scaled by the expected variance [Felsenstein:1985ua]. The expected variance is principally determined by the lengths of the branches leading from the node to these daughters. The shorter the total length of the two branches leading to dauter nodes, the larger the contrast for a given observed difference. This is because the same difference across a shorter total branch length indicates a greater rate of evolutionary change. The expected variance of contrasts for speciation nodes is constrained by the branch lengths on the species tree, but the expected variance of contrasts for duplications has a much wider range (Figure XXAscertainment c). This could lead to biases if the lengths of branches that descend from duplication nodes tend to be overestimated. We thereforeexamined only the contrasts that fell within in the range of expected variance seen for speciation contrasts, excluding duplication contrasts that fall outside of this range. This reanalysis does not reject the ortholog conjecture either (Figure XXAscertainment d), indicating that branch length bias is not responsible for the result.

@@

The reason that paralog pairs have Tau values that are more different than ortholog pairs is that they tend to be more distantly related, not because there are different rates of Tau evolution following duplication and speciation events. This means that having information on whether two genes are orthologs or paralogs provides little added information about expression beyond knowing how distantly related the two genes are. This has several implications. It is an example of the limitations of the concepts of orthology and paralogy [@Dunn:2016er]. These terms can have straightforward meaning in small gene trees with simple duplication/speciation histories, but the utility of the terms breaks down on larger more complex gene trees. Orthology and paralogy are annotations on the tips of the tree that are derived from the infered structure of the tree and annotations of duplication and speciation at internal tree nodes. In this sense, orthology and paralogy are statements about the internals of the tree that are distilled into statements at the tips of the tree. Much is lost in the process, though. In this case, KMRR [@KryuchkovaMostacci:2016iw] relied on orthology and paralogy annotations that are derived from the infered duplication and speciation history on gene phylogenies [@Herrero:2016jj]. These are the same pre-computed trees we use in our phylogenetic re-analyses of their study. For most questions it is much more direct to focus on the structure of the tree and the inferred processes within the tree, such as which internal nodes are duplication or speciation events and how much change occurs along the branches.

@@

@@

How changes in animal development relate to the evolution of animal diversity is a major question in evolutionary developmental biology (EvoDevo). To address this topic, Levin *et al.* [@Levin:2016bp] analyzed gene expression through the course of embryonic development for ten animal species, each from a different phylum. As their title ("The mid-developmental transition and the evolution of animal body plans") indicates, they arrived at two major conclusions. First, animal development is characterized by a well-defined mid-developmental transition. Second, this transition helps explain the evolution of features observed among distantly related animals. Specifically, they concluded that animals from different phyla exhibit an "inverse hourglass" model for the evolution of gene expression, where there is more evolutionary variance in gene expression at a mid phase of development than there is at early and late phases. Closely related animals have previously been described as having an hourglass model of gene expression, where evolutionary variance in expression is greater early and late in development than at the midpoint of development [@Kalinka:2010dw; @DomazetLoso:2010cn]. Levin *et al.* conclude that this contrast between distantly and closely related animals provides biological justification for the concept of phyla and may provide a long-sought operational definition of phyla. We previously described some concerns with their interpretations of these results [@Hejnol:2016hw].

Rather than support the inverse hourglass as a general pattern among distantly related species, our reanalyses of Levin *et al.* [@Levin:2016bp] indicate their existing data support the opposite conclusion - that the inverse hourglass is due to large differences specific to a single lineage, the ctenophore (comb jelly). Better understanding these differences will be of interest to future work, especially in light of current uncertainty regarding the phylogenetic placement of ctenophores and their unique biology [@Martindale:2015ve, @Dunn:2015ij].

To understand our concerns with their analyses, it is helpful to first outline Levin *et al.*'s published methods and results. Although there are mature methods and tools for statistical analysis of character evolution [@Garamszegi:2014ew] and gene expression [@Robinson:2009cw], Levin *et al.* [@Levin:2016bp] drew on neither of these very active areas of research and instead applied an *ad hoc* pairwise comparison of orthologous gene expression between species. They characterized each gene in each species as having expression that peaks in early, mid, or late temporal phase of development while the goodness of fit to these patterns was not considered, and alternative patterns were not evaluated. For each species pair, they then identified the orthologs shared by these species (shared orthologs vary from pair to pair). They then calculated a similarity score for each temporal phase for each species pair based on the fraction of genes that exhibited the same patterns in each species. The distributions of similarity scores are plotted in their [Figure 4d](http://www.nature.com/nature/journal/v531/n7596/fig_tab/nature16994_F4.html), and their Kolmogorov–Smirnov (KS) tests indicated that the early distribution and late distribution were each significantly different from mid distribution (P < 10<sup>-6</sup> and P < 10<sup>-12</sup>, respectively). This is the support they presented for the inverse hourglass model.

We began by examining the matrix of pairwise comparisons that their KS tests and Figure 4d are based on. As this figure shows (regenerated here as our Figure XXLevin a), there is not a strong distinction between these temporal phases - the distributions for the early, mid, and late distributions overlap considerably. The medians are close in value, and the range of mid phase includes the entire range of the early phase. Further inspection (Figure XXLevin a) reveals that the mid phase distribution has several outliers with very low similarity scores. We found that all five of the lowest values in the mid phase distribution are for pairwise comparisons that include the ctenophore (Figure XXLevin a). When the nine pairwise comparisons that include the ctenophore are removed, the differences between the early and mid phase distributions are greatly reduced (Figure XXLevin b).

We also found several problems with the statistical tests that were used to evaluate the inverse hourglass hypothesis. First, in the published analyses every data point was included twice because both reciprocal comparisons (which have the same values) were retained. For example, there is both a nematode to arthropod comparison and an arthropod to nematode comparison. As a consequence, there are 90 entries for the 45 pairwise comparisons, and by doubling the data the significance of the result appears much stronger than it actually is. After removing the duplicate values, the p values are far less significant, 0.002 for the early-mid comparison and on the order of 10<sup>-6</sup> for early-late. Second, the test they used (KS test) is not appropriate for the hypothesis they seek to evaluate. The KS test does not just evaluate whether one distribution is greater than the other, it also tests whether the shape of the distributions are the same. In addition, the samples in this dataset are matched (*i.e.*, for each pairwise comparison there is a early, mid, and late expression value), which the KS test does not take into account. The Wilcoxon test is instead appropriate in this case. After removing the duplicate scores, removing the ctenophore, and applying the Wilcoxon test, there is no significant difference between the early phase and mid phase distributions (P = 0.1428 for the early-mid comparison and P < 10<sup>-5</sup> for the late-mid comparison), and the inverse hourglass turns into a wine bottle (Figure XXLevin b).

Our results highlight the importance of explicitly accounting for phylogenetic relationships when studying character evolution, including developmental [@Telford:2003tsa] and functional genomic traits [@Hejnol:2015gr]. This is particularly true for evolutionary analyses of quantitative gene expression [@Dunn:2013cpa]. Pairwise comparisons result in the same evolutionary changes being counted multiple times in each species pair, as for the ctenophore comparisons that were mistaken for a global pattern here. This is a well understood property of pairwise comparisons that has been specifically addressed by phylogenetic comparative methods [@Felsenstein:1985ua]. This particular case illustrates the shortcomings of *ad hoc* rank-based methods [@DomazetLoso:2010cn] and pairwise comparisons [@Levin:2016bp] for the study of the evolution of gene expression. Phylogenetic analyses of character evolution provide not only a way to avoid these problems, but a richer context for interpreting the biological implications of the results.

While we do demonstrate a common problem with pairwise comparisons in the Levin *et al.* analysis, we did not perform a phylogenetic reanalysis of this study, as we did for the KMRR study. This is because the similarity metric computed in the pairwise comparisons of Levin *et al.* are based on different genes for different species pairs. This means that there is not a trait that can be mapped onto the tree, as KMRR's Tau can be. A full phylogenetic reanalysis would of course be possible be using upstream analysis products to re-derive a new expression summary statistic that was suitable for mapping onto a phylogenetic tree.

### Comparative methods in functional genomics

Some of the most widely used phylogenetic comparative methods [1, 5] are directly applicable to CFG data. There are also unique challenges that will need to be addressed at this new intersection, though. One of the greatest challenges is that most phylogenetic comparative methods have been developed to address problems with many more species (e.g., dozens or more) relative to the number of traits being examined (e.g., 2-5). In CFG analyses, there are often far fewer species (e.g., 3-10) because adding taxa is still expensive, but tens of thousands of traits. This creates challenges as the resulting covariance matrices are singular and, if not treated appropriately, imply many false correlations that are artifacts of project design. We outlined these challenges and potential solutions in the context of gene expression [31].

It is not enough to demonstrate the impact of phylogenetic methods on CFG analyses, these methods need to be made accessible to the functional genomics community so they can be widely applied.

Most early phylogenetic comparative methods attempted to remove evolutionary signal to correct statistical tests for correlations between traits, while more recent methods tend to focus on testing hypotheses of evolutionary processes [34]. The application of this newer focus to pCFG provides an exciting opportunity to address long standing questions of broad interest, including the order of changes in functional genomics traits and shifts in rates of evolution of one functional genomics trait following changes in another trait.

A small number of CFG studies have employed some phylogenetic comparative approaches [18–22]. For instance, a phylogenetic ANOVA [21] of the evolution of gene expression improves statistical power and drastically reduces the rate of false positives relative to pairwise approaches.

The other issue with the pairwise comparisons is that, except at the tips, they summarize changes along many branches in the phylogeny. Two paralogs that diverged from a duplication event deep in the tree may have many subsequent duplication and speciation events, and changes along all these branches will impact the final pairwise comparison. Phylogenetic methods have the advantage of isolating the changes under consideration (Fig. 1b). The phylogenetic methods avoid diluting the change that occurs along the branches that follow the node in question with changes along all subsequent branches. There may still be missing speciation events, due to extinction and incomplete taxon sampling, and missing duplication events, due to gene loss, but these omissions affect both methods.

@@

The fact that the first two CFG studies we reanalyzed show serious problems with pairwise comparisons indicates that there likely to be similar problems in other CFG studies, and that future CFG studies will be compromised if they continue to use pairwise methods.

Studies of evolutionary functional genomics should not be focused on the tips of the tree, they should explicitly delve into the tree itself.

@@

Each of these gene trees includes includes homologous sequences derived from a common ancestral sequence. The tip nodes of a gene tree can include both orthologs and paralogs, and each internal node is an inferred speciation or duplication event. Prior to phylogenetic analyses of gene expression, each tree was first pruned to include only tips for which expression data were available. The speciation nodes in the tree were then time calibrated so that branch lengths were consistent across trees. We then calculated the magnitude of the independent contrast for differential expression at each internal node in each of these trees. The independent contrast is the amount of change in expression observed along the branches that arise from the node, scaled by the expected variance [Felsenstein:1985ua]. The expected variance is determined mostly by the length of the branches, but can also be influenced by uncertainty regarding the reconstruction of differential expression at internal nodes. We took the absolute value of these contrasts, since the sign depends only on the arbitrary ordering of the nodes in the trees.

>>>>

A rich theoretical and statistical methodology of phylogenetic comparative methods have been developed over the last three decades to address the challenges and opportunities of comparisons across species [@Felsenstein:1985ua; @Grafen:1989um; @Chamberlain:2012vx; @Garamszegi:2014ew; @Pagel:1999gc]. These challenges include non-independence of observations across species -- more closely related species will tend to be more similar to each other than distantly related species. Phylogenetic comparative methods have largely been applied to the evolution of morphological and ecological traits, but are just as relevant to CFG. Most CFG studies have abstained from phylogenetic approaches and instead heavily rely on multiple pairwise comparisons (Fig. 1A) between contemporary traits that do not account for the fact that evolutionary relationships explain much of the structure of variation across species. This leaves CFG studies susceptible to serious statistical errors and is a missed opportunity to ask questions that are only accessible in an explicit phylogenetic context.

@@

The KMRR study [@KryuchkovaMostacci:2016iw] sought to test the ortholog conjecture. The ortholog conjecture [@Nehrt:2011bm] is the proposition that orthologs (genes that diverged from each other due to a speciation event) have more similar attributes than do paralogs (genes that diverged from each other due to a gene duplication event). The ortholog conjecture has important biological and technical implications. It shapes our understanding of the functional diversity of gene families. It is used to relate findings from well-studied genes to related genes that have not been investigated in detail. It can be applied to any trait of genes, from gene sequence to biochemical properties to expression. While the ortholog conjecture describes a specific pattern of functional diversity across genes, it is also articulated as a hypothesis about the process of evolution-- that there is greater evolutionary change in gene attributes following a duplication event than a speciation event.

@@

The independent contrast across a node is the amount of change observed between the daughter notes, scaled by the expected variance [@Felsenstein:1985ua]. The expected variance is principally determined by the lengths of the branches leading from the node to these daughters. The shorter the total length of the two branches leading to daughter nodes, the larger the contrast for a given observed difference. This is because the same difference across a shorter total branch length indicates a greater rate of evolutionary change. The expected variance of contrasts for speciation nodes is constrained by the branch lengths on the species tree, but the expected variance of contrasts for duplications has a much wider range (Figure XXAscertainment c). This could lead to biases if the lengths of branches that descend from duplication nodes tend to be overestimated. We therefore examined only the contrasts that fell within in the range of expected variance seen for speciation contrasts, excluding duplication contrasts that fall outside of this range. This reanalysis does not reject the ortholog conjecture either (Figure XXAscertainment d), indicating that branch length bias is not responsible for the result.

@@

The reason that paralog pairs have Tau values that are more different than ortholog pairs is that they tend to be more distantly related, not because there are different rates of Tau evolution following duplication and speciation events. This means that having information on whether two genes are orthologs or paralogs provides little added information about expression beyond knowing how distantly related the two genes are. This has several implications. It is an example of the limitations of the concepts of orthology and paralogy [@Dunn:2016er]. These terms can have straightforward meaning in small gene trees with simple duplication/speciation histories, but the utility of the terms breaks down on larger more complex gene trees. Orthology and paralogy are annotations on the tips of the tree that are derived from the inferred structure of the tree and annotations of duplication and speciation at internal tree nodes. In this sense, orthology and paralogy are statements about the internals of the tree that are distilled into statements at the tips of the tree. Much is lost in the process, though. In this case, KMRR [@KryuchkovaMostacci:2016iw] relied on orthology and paralogy annotations that are derived from the inferred duplication and speciation history on gene phylogenies [@Herrero:2016jj]. These are the same pre-computed trees we use in our phylogenetic re-analyses of their study. For most questions it is much more direct to focus on the structure of the tree and the inferred processes within the tree, such as which internal nodes are duplication or speciation events and how much change occurs along the branches.

#### Implications for the ortholog conjecture

@@

#### Original pairwise analyses of developmental gene expression

@@

How changes in animal development relate to the evolution of animal diversity is a major question in evolutionary developmental biology (EvoDevo). To address this topic, Levin *et al.* [@Levin:2016bp] analyzed gene expression through the course of embryonic development for ten animal species, each from a different phylum. As their title ("The mid-developmental transition and the evolution of animal body plans") indicates, they arrived at two major conclusions. First, animal development is characterized by a well-defined mid-developmental transition that marks the transition from an early phase of gene expression to a late stage of gene expression. Second, this transition helps explain the evolution of features observed among distantly related animals. Specifically, they concluded that animals from different phyla exhibit an "inverse hourglass" model for the evolution of gene expression, where there is more evolutionary variance in gene expression at a mid phase of development than there is at early and late phases. Closely related animals have previously been described as having an hourglass model of gene expression, where evolutionary variance in expression is greater early and late in development than at the midpoint of development [@Kalinka:2010dw; @DomazetLoso:2010cn]. Levin *et al.* conclude that this contrast between distantly and closely related animals provides biological justification for the concept of phyla and may provide a definition of phyla.

Although there are mature methods and tools for statistical analysis of character evolution [@Garamszegi:2014ew] and gene expression [@Robinson:2009cw], Levin *et al.* [@Levin:2016bp] drew on neither of these very active areas of research and instead applied an *ad hoc* pairwise comparison of orthologous gene expression between species. They characterized each gene in each species as having expression that peaks in early, mid, or late temporal phase of development. The goodness of fit to these patterns was not considered, and alternative patterns, such as the possibility that there are more phases of expression or no distinct phases of expression, were not evaluated. For each species pair, they identified the orthologs shared by these species (shared orthologs vary from pair to pair). They then calculated a similarity score for each temporal phase for each species pair based on the fraction of genes that exhibited the same patterns in each species. The distributions of similarity scores are plotted in their [Figure 4d](http://www.nature.com/nature/journal/v531/n7596/fig_tab/nature16994_F4.html), and their Kolmogorov–Smirnov (KS) tests indicated that the early distribution and late distribution were each significantly different from mid distribution (P < 10<sup>-6</sup> and P < 10<sup>-12</sup>, respectively). This is the support they presented for the inverse hourglass model.

We previously described some concerns with their interpretations of these results [@Hejnol:2016hw]. The main interpretation issues we identified with their interpretation is that because they only sampled one species per phylum, there is no way to tell if observed differences are unique to the branches that gave rise to phyla or occurred along the many other branches in the tree. By the stated logic this study could just as easily be a test of unique expression differences between classes, orders, or families, since each of the sampled animals come from a different one of these groups. It is far more likely that changes between such distantly related taxa do not correspond to taxonomic ranks in any way.  Phyla were invented as taxonomic convenience to refer to animal groups of unknown relationship to each other, so they are a statement of historical phylogenetic uncertainty by humans rather than an intrinsic biological phenomenon to be explained.

#### Reexamination of pairwise comparisons

Here we examine the support for the inverse hourglass model itself. We began by examining the matrix of pairwise comparisons that their KS tests and Figure 4d are based on. As this figure shows (regenerated here as our Figure XXLevin a), there is not a strong distinction between these temporal phases - the distributions for the early, mid, and late distributions overlap considerably. The medians are close in value, and the range of mid phase includes the entire range of the early phase. Further inspection (Figure XXLevin a) reveals that the mid phase distribution has several outliers with very low similarity scores. We found that all five of the lowest values in the mid phase distribution (Figure XXLevin a) are for pairwise comparisons that include the ctenophore (comb jelly). When the nine pairwise comparisons that include the ctenophore are removed, the differences between the early and mid phase distributions are greatly reduced (Figure XXLevin b).

We also found several problems with the statistical tests that were used to evaluate the inverse hourglass hypothesis. First, in the published analyses every data point was included twice because both reciprocal comparisons (which have the same values) were retained. For example, there is both a nematode to arthropod comparison and an arthropod to nematode comparison. As a consequence, there are 90 entries for the 45 pairwise comparisons, and by doubling the data the significance of the result appears much stronger than it actually is. After removing the duplicate values, the p values are far less significant, 0.002 for the early-mid comparison and on the order of 10<sup>-6</sup> for early-late. Second, the test they used (KS test) is not appropriate for the hypothesis they seek to evaluate. The KS test does not just evaluate whether one distribution is greater than the other, it also tests whether the shape of the distributions are the same. In addition, the samples in this dataset are matched (*i.e.*, for each pairwise comparison there is a early, mid, and late expression value), which the KS test does not take into account. The Wilcoxon test is instead appropriate in this case. After removing the duplicate scores, removing the ctenophore, and applying the Wilcoxon test, there is no significant difference between the early phase and mid phase distributions (P = 0.1428 for the early-mid comparison and P < 10<sup>-5</sup> for the late-mid comparison), and the inverse hourglass turns into a wine bottle (Figure XXLevin b). Rather than support the inverse hourglass as a general pattern among distantly related species, our reanalyses of Levin *et al.* [@Levin:2016bp] comes to a very conclusion - the greater variance at the midpoint of development is specific to a single lineage, the ctenophore.

Our results highlight the importance of explicitly accounting for phylogenetic relationships when studying character evolution, including developmental [@Telford:2003tsa] and functional genomic traits [@Hejnol:2015gr]. This is particularly true for evolutionary analyses of quantitative gene expression [@Dunn:2013cpa]. Pairwise comparisons result in the same evolutionary changes being counted multiple times in each species pair, as for the ctenophore comparisons that were mistaken for a global pattern here. This is a well understood property of pairwise comparisons that has been specifically addressed by phylogenetic comparative methods [@Felsenstein:1985ua]. This particular case illustrates the shortcomings of *ad hoc* rank-based methods [@DomazetLoso:2010cn] and pairwise comparisons [@Levin:2016bp] for the study of the evolution of gene expression.

While we do demonstrate a common problem with pairwise comparisons in the Levin *et al.* analysis, we did not perform a phylogenetic reanalysis of this study, as we did for the KMRR study. This is because the similarity metric computed in the pairwise comparisons of Levin *et al.* are based on different genes for different species pairs. This means that they do not generate a trait that can be evaluated across a phylogeny, as KMRR's Tau can be. A full phylogenetic reanalysis would be possible by using upstream analysis products to re-derive new expression summary statistics suitable for phylogenetic comparative methods.

### Phylogenetic comparative methods in functional genomics

Some of the most widely used phylogenetic comparative methods [@Felsenstein:1985ua; @Grafen:1989um] are directly applicable to CFG data. There are also unique challenges that will need to be addressed at this new interdisciplinary interface, though. One of the greatest challenges is that most phylogenetic comparative methods have been developed to address problems with many more species (e.g., dozens or more) relative to the number of traits being examined. In CFG analyses, there are often far fewer species because adding taxa is still expensive, but tens of thousands of traits. This creates statistical challenges as the resulting covariance matrices are singular and, if not treated appropriately, imply many false correlations that are artifacts of project design. We outlined these challenges and potential solutions in the context of gene expression [@Dunn:2013cpa]. In addition to methods development at this interface, there is a critical need for software tools that make it simpler to integrate existing functional genomic and phylogenetic comparative packages.

Recent advances in phylogenetic comparative methods are particularly well suited to addressing questions about the evolution of functional genomic traits. Most early phylogenetic comparative methods attempted to remove evolutionary signal to correct statistical tests for correlations between traits, while more recent methods tend to focus on testing hypotheses of evolutionary processes [@Pennell:2013cw]. The application of this newer focus to functional genomics provides an exciting opportunity to address long standing questions of broad interest, including the order of changes in functional genomics traits and shifts in rates of evolution of one functional genomics trait following changes in another trait.

We are not the first to apply phylogenetic comparative methods to functional genomic data. While the vast majority of comparative functional genomics studies have used pairwise similarity methods, a small number of CFG studies have employed phylogenetic comparative approaches [@Oakley:2005ky; @Eng:2009gp; @Chang:2014hq]. For instance, a phylogenetic ANOVA [@Rohlfs:2015bd] of the evolution of gene expression improves statistical power and drastically reduces the rate of false positives relative to pairwise approaches.

Addressing the statistical non-independence of pairwise comparisons is not the only advantage of using phylogenetic comparative methods for functional genomics analyses. Another problem with the pairwise comparisons is that, except at the tips, they summarize changes along many branches in the phylogeny. Two paralogs that diverged from a duplication event deep in the tree may have many subsequent duplication and speciation events, and changes along all these branches will impact the final pairwise comparison. Phylogenetic methods have the advantage of isolating the changes under consideration (Fig. 1b). The phylogenetic methods avoid diluting the change that occurs along the branches that follow the node in question with changes along all subsequent branches. There may still be missing speciation events, due to extinction and incomplete taxon sampling, and missing duplication events, due to gene loss, but these omissions affect both methods.

@@

The fact that the first two CFG studies we reanalyzed show serious problems with pairwise comparisons indicates that there likely to be similar problems in other CFG studies, and that future CFG studies will be compromised if they continue to use pairwise methods. Studies of evolutionary functional genomics should not be focused on the tips of the tree, they should explicitly delve into the tree itself.

@@

Each of these gene trees includes includes homologous sequences derived from a common ancestral sequence. The tip nodes of a gene tree can include both orthologs and paralogs, and each internal node is an inferred speciation or duplication event. Prior to phylogenetic analyses of gene expression, each tree was first pruned to include only tips for which expression data were available. The speciation nodes in the tree were then time calibrated so that branch lengths were consistent across trees. We then calculated the magnitude of the independent contrast for differential expression at each internal node in each of these trees.
